risk-perception

### 2019-10-23 risk-percption.Rmd
rearrange opener with more explanatory plots

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(zoo)
```

```{r}
test <- tibble(agent = 1:5,
       alpha = 1.5, #runif(5),
       beta = 1.5) %>%
  mutate(mu = alpha / (alpha + beta))

@@

The basic risk assessment agorithm. Let's look at the risk of a drought year being 0.2. In a wet year you get 3 wheat and 2 barley, and in a dry year you get 0 wheat and 1 barley. What does that look like?

```{r}
wheat_dry <- 0
wheat_wet <- 1.6
barley_dry <- 0.93
barley_wet <- 1.18

calc_payoff <- function(wheat_prop, risk) {
  barley <- (1 - wheat_prop) * (barley_dry * risk + barley_wet * (1 - risk))
  wheat <- wheat_prop * (wheat_dry * risk + wheat_wet * (1 - risk))
  barley + wheat
}

tibble(wheat_prop)
```

calc_ratio <- function(risk) {
  (risk * barley_dry + (risk - 1) * barley_wet) / (risk * (barley_dry - wheat_dry) + wheat_wet * -1 * risk + wheat_wet + barley_wet * (risk - 1))
}
calc_ratio(.2)
```

So at between 3 in 10 and 4 in 10 risk of drought, a maximizer will switchfrom wheat to barley.
```{r}
test_dat <- expand_grid(wheat_prop = (0:10)/10,
       risk = seq(0, 1, 0.2)) %>%
  mutate(util_wet = (1 - risk) * (wheat_wet * wheat_prop + barley_wet * (1 - wheat_prop)),
         util_dry = risk * (wheat_dry * wheat_prop + barley_dry * (1 - wheat_prop)),
         util_wheat = wheat_prop * (wheat_wet * (1 - risk) + wheat_dry *  risk),
         util_barley = (1 - wheat_prop) * (barley_wet* (1 - risk)+ barley_dry * risk),
         payoff = calc_payoff(wheat_prop, risk))

test_dat %>%
  group_by(risk) %>%
  filter(payoff == max(payoff)) %>%
  ggplot(aes(risk, wheat_prop)) + geom_point()

test_dat %>%
  select(-(util_dry:util_wet)) %>%
  gather(type, util, util_wheat:util_barley) %>%
    ggplot(aes(wheat_prop, util, color = type)) + geom_line() + facet_wrap(~risk)
test_dat %>%
  select(-(util_wheat:util_barley)) %>%
  gather(type, util, util_dry:util_wet) %>%
    ggplot(aes(wheat_prop, util, color = type)) + geom_line() + facet_wrap(~risk)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lpSolve)

#library(zoo)
```

Start with a population of agents with the same subjective beliefs about drought risk.

```{r}
agents <- tibble(id = 1:5, 
                 alpha = 1.5,
                 beta = 1.5,
                 risk = alpha / (alpha + beta))
```

```{r}
agents
```

And payoffs for different crop types in different years.
```{r}
wheat_dry <- 0
wheat_wet <- 1.6
barley_dry <- 0.93
barley_wet <- 1.18

wheat_dry <- .2
wheat_wet <- .4
barley_dry <- 0.6
barley_wet <- .3
```

The expected yield is thus calculated as.
```{r}
expected_utility <- function(risk) {
  risk * (wheat_dry + barley_dry) + (1 - risk) * (wheat_wet + barley_wet) 
}
```

```{r}
expected_utility(0.2) # assuming a drought risk of 0.2
```

Given these payoffs, what's the point of indifference between the two crop types?

```{r, echo = FALSE}
plt_dat <- tibble(risk = seq(0, 1, .1),
       barley_exp = risk * barley_dry + (1 - risk) * barley_wet,
      wheat_exp = risk * wheat_dry + (1 - risk) * wheat_wet) %>%
  gather(type, utility, barley_exp:wheat_exp)
```

The pure maximin strategy is to plant barley:
```{r}
max(wheat_dry, barley_dry)
```

```{r, echo = FALSE}
indiff <- tibble(risk = (barley_wet - wheat_wet) / (wheat_dry - wheat_wet - barley_dry + barley_wet),
                 utility = risk * wheat_dry + (1 - risk) * wheat_wet)

plt_dat %>%
  ggplot(aes(risk, utility)) +
  geom_hline(yintercept = indiff$utility, linetype = 2)+
  geom_vline(xintercept = indiff$risk, linetype = 2) +
  geom_line(aes(color = type), size = 1.2) +
  geom_point(data = indiff, color = 'black', size = 3) +
  geom_text(data = indiff, label = '(0.3, 1.1)', nudge_x = 0.1, nudge_y = 0.1) +
  theme_classic()
```

The maximin strategy here is to plant barley all the time, but this entails a huge opportunity cost over planting wheat or a mix of the two

```{r, echo = FALSE}
plt_dat2 <- tibble(wheat_prop = seq(0, 1, .1),
       dry_exp = wheat_prop * wheat_dry + (1 - wheat_prop) * barley_dry,
      wet_exp = wheat_prop * wheat_wet + (1 - wheat_prop) * barley_wet) %>%
  gather(type, utility, dry_exp:wet_exp)
```

So the intersection point of the next plot is the mixed strategy equilbrium(indifference point) in a risk-free environment

```{r, echo = FALSE}
nash_eq <- tibble(wheat_prop = abs(barley_dry - barley_wet) / (abs(barley_dry - barley_wet) + abs(wheat_dry - wheat_wet)),
                  utility = wheat_prop * (wheat_dry + wheat_wet))

plt_dat2 %>%
  ggplot(aes(wheat_prop, utility)) +
  geom_hline(yintercept = nash_eq$utility, linetype = 2)+
  geom_vline(xintercept = nash_eq$wheat_prop, linetype = 2) +
  geom_line(aes(color = type), size = 1.2) +
  geom_point(data = nash_eq, color = 'black', size = 3) +
  geom_text(data = nash_eq, 
            label = paste0('(', nash_eq$wheat_prop, ', ', nash_eq$utility, ')'),
            nudge_x = 0.1, nudge_y = 0.05) +
  theme_classic()

when you sample from the expected value, use that to choose a strategy. but if you sample lots, use that to create a mixed streateg

And the subjective expected yields for the agents are thus.
```{r}
agents %>%
  mutate(risk = runif(n())) %>%
  mutate(yield_exp = expected_utility(risk),
         barley_exp = risk * barley_dry + (1 - risk) * barley_wet,
         wheat_exp = risk * wheat_dry + (1 - risk) * wheat_wet,
         favorite = case_when(barley_exp == wheat_exp ~ 'either',
                              barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat')) 
```

```{r}
calc_ratio <- function(risk) {
  pmax(0, (risk * barley_dry + (risk - 1) * barley_wet) / (risk * (barley_dry - wheat_dry) + wheat_wet * -1 * risk + wheat_wet + barley_wet * (risk - 1)))
}
```

Now create a population of agents with different subjective beliefs about the drought risk.
```{r}

@@

The basic risk assessment algorithm. Let's look at the risk of a drought year being 0.2. In a wet year you get 3 wheat and 2 barley, and in a dry year you get 0 wheat and 1 barley. What does that look like?

maximax
```{r}
risk <- 0.9
wheat_exp <- wheat_wet * (1 - risk) + wheat_dry * risk
barley_exp <- barley_wet * (1 - risk) + barley_dry * risk
obj.fun <- c(wheat_exp, barley_exp)

constr=matrix(c(1,1,1,0,0,1, wheat_exp, barley_exp),ncol = 2, byrow = TRUE) #Loading the constraints
constr.dir=c("=",">=",">=", ">=")
constr.rhs=c(1,0,0, .5)

mod=lp("min",obj.fun,constr,constr.dir,constr.rhs) #Using lp() to solve our problem
mod$solution
```

```{r}
calc_payoff <- function(x, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_wet; barley_yield <- barley_wet

  x %>%
    mutate(yield = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop))
}

update_prop <- function(x, aspiration) {
  x %>%
    mutate(new_prop = round(runif(n()), 2),
           wheat_prop = if_else(yield > aspiration, wheat_prop, new_prop)) %>%
    select(-new_prop)
}

test3 <- tibble(agent = 1:1000) %>%
  mutate(wheat_prop = round(runif(n()), 2)) %>%
  calc_payoff(drought = TRUE)
test3 %>% update_prop(aspiration = .6)

accumulate((runif(25) < .3), ~calc_payoff(.x, drought = .y) %>% update_prop(aspiration = .5), .init = test3) %>% bind_rows(.id = 'time') %>% ggplot(aes(as.numeric(time), yield, group = agent)) + geom_line(alpha = .2)

```{r}
library(ggridges)
accumulate((runif(25) < .2), ~calc_payoff(.x, drought = .y) %>% update_prop(aspiration = .6), .init = test3) %>% bind_rows(.id = 'time') %>% ggplot(aes(wheat_prop, y = as.numeric(time), group = as.numeric(time))) +
  geom_density_ridges() +
  theme_minimal()
```

So at between 3 in 10 and 4 in 10 risk of drought, a maximizer will switchfrom wheat to barley.

```{r}
test_dat <- expand_grid(wheat_prop = (0:10)/10,
       risk = seq(0, 1, 0.1)) %>%
  mutate(util_wet = (1 - risk) * (wheat_wet * wheat_prop + barley_wet * (1 - wheat_prop)),
         util_dry = risk * (wheat_dry * wheat_prop + barley_dry * (1 - wheat_prop)),
         util_wheat = wheat_prop * (wheat_wet * (1 - risk) + wheat_dry *  risk),
         util_barley = (1 - wheat_prop) * (barley_wet* (1 - risk) + barley_dry * risk),
         util_payoff = util_dry + util_wet)
```

```{r}
test_dat %>%
  select(-(util_wheat:util_barley)) %>%
  gather(type, util, c(util_dry, util_wet, util_payoff)) %>%
    ggplot(aes(wheat_prop, util, color = type)) + geom_line() + facet_wrap(~risk) + geom_hline(yintercept = .5, alpha = .3, linetype = 2) + geom_vline(xintercept = .14, alpha = 0.3,linetype = 2)
```
```{r}
test_dat %>%
  select(-(util_wheat:util_barley)) %>%
  mutate(util_min = pmin(util_wet, util_dry))%>%
    ggplot(aes(util_payoff, util_min)) + geom_line() + facet_wrap(~risk) + geom_hline(yintercept = .3, alpha = .3, linetype = 2)
```

```{r}
test_dat %>%
  group_by(risk) %>%
  filter(payoff == max(payoff)) %>%
  ggplot(aes(risk, wheat_prop)) + geom_point()

test_dat %>%
  select(-(util_dry:util_wet)) %>%
  gather(type, util, util_wheat:util_barley) %>%
    ggplot(aes(wheat_prop, util, color = type)) + geom_line() + facet_wrap(~risk) + geom_vline(xintercept = .14)
